

---
title: "Prelab_demo_Clean_Analyze_Death_Certificates"
author: "Wells"
date: "2025-11-17"
output: html_document
---

#Name: Nicole Ramos & Bex Heimbrock & Raphael Romero Ruiz
#Date: 11-17-2025
#Assignment: Pre-lab_clean_death_certificates


This exercise will involve cleaning, tabulating and mapping the death certificate data from Cheltenham and Jessup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages("htmlwidgets")
install.packages("leaflet")
install.packages("sf")
install.packages("googlesheets4")
library(tidyverse)
library(rio)
library(janitor)
library(htmlwidgets)
library(leaflet)
library(sf)
library(googlesheets4)
library(tidycensus)
```

### Import Extracted Death Certificate Data
```{r}

gs4_deauth()  
dc <- read_sheet("https://docs.google.com/spreadsheets/d/13CthhhfLHiPZqr1P67SCa9BVUyb5Rl6OBhUwvqawDEk/edit?usp=sharing", sheet = "responses_copy") |> 
    select(-Duplicates) |> 
  clean_names()


```



## Mapping place of origin

```{r}

dc |> 
  mutate(origin = birth_place_city_or_town_and_state_write_it_as_bowie_md) |>
  count(origin) |> 
  arrange(desc(n))


```

### Massive clean origin codeblock
```{r}

dc <- dc |> 
  mutate(origin = str_squish(birth_place_city_or_town_and_state_write_it_as_bowie_md)) |> 
  mutate(origin = str_remove_all(origin, "[?|()]"))|>
  mutate(county_origin = case_when(
         str_detect(origin, "(?i)^md\\.?$") ~ "Maryland", 
         str_detect(origin, "(?i)Frederick, MD") ~ "Frederick County",
         str_detect(origin, "(?i)baltimore county") ~ "Baltimore County",  
         str_detect(origin, "(?i)baltimore") ~ "Baltimore city",
         str_detect(origin, "(?i)Cheltenham") ~ "Prince George's County",
        str_detect(origin, "(?i)Church Hill") ~ "Queen Anne's County",
        str_detect(origin, "(?i)Dunkirk, Md.") ~ "Calvert County",
        str_detect(origin, "(?i)Easton Md") ~ "Talbot County",
        str_detect(origin, "(?i)Easton, Md") ~ "Talbot County",
        str_detect(origin, "(?i)Elkton, Md.") ~ "Cecil County",
        str_detect(origin, "(?i)Ellicott City MD") ~ "Howard County",
        str_detect(origin, "(?i)Harford County") ~ "Harford County",
        str_detect(origin, "(?i)Illegible") ~ "Unknown",
        str_detect(origin, "(?i)Jessup") ~ "Anne Arundel County", 
            str_detect(origin, "(?i)Md. Residence Oakland, MD.") ~ "Garrett County",
        str_detect(origin, "(?i)Md. former residence Pocomoke, Md.") ~ "Worcester County",
        str_detect(origin, "(?i)Not known") ~ "Unknown",
        str_detect(origin, "Pinna former residence Jamestown, Md.") ~ "Stutsman County",
        is.na(origin) ~ "Unknown",
               TRUE ~ origin
         )) |> 
      mutate(county_origin = str_remove_all(county_origin, ", Md."))
```

```{r}
dc |> 
  count(county_origin) |> 
  arrange(desc(n))


```

# Map
### Retrieve county polygons

```{r echo=FALSE}
census_api_key("", install=FALSE)
```

```{r}

map <- get_acs(geography = "county", 
              variables = c(number_households = "B19001_001", median_income = "B19013_001"), 
               state = "MD", 
               year = 2020, geometry = TRUE) %>% 
  mutate(NAME = str_remove_all(NAME, ", Maryland")) |> 
  select(GEOID, NAME) |> 
  distinct()
```

### Join geocoordinates with death certificates

```{r}

dc_map <- dc |>  
 inner_join(map, by=c("county_origin"="NAME")) 

```

### Data to a coordinate reference system
transforms the coordinate reference system (CRS) of a spatial object. Need to do this when plotting data on web maps and need to use latitude/logitude
```{r}

dc_map <- dc_map |> 
  st_as_sf() |> 
  sf::st_transform(4326) |> 
  add_count(county_origin, name = "county_count")
```



```{r}

# Use colorFactor for categorical data
pal <- colorFactor(
  palette = "viridis",  # or "Set1", "Dark2", etc for categories
  domain = dc_map$county_count
)


leafMap <- leaflet(data = dc_map) %>%
  addProviderTiles(providers$Esri.WorldStreetMap) %>%
  addPolygons(
    color = ~pal(county_count),
    weight = 2.5,
    fillOpacity = 0.5,
    label = ~paste0("Count of inmates from ", county_origin, ": ", county_count)
  ) 
  
leafMap
```


**Your Turn**
Split up into groups. Each group will tackle various towns, counties.
Coordinate so you don't overlap.

Clean and categorize more counties. Examine the data and modify the Massive clean origin codeblock, add more to the county_origin

```{r}
dc |> 
  count(county_origin) |> 
  arrange(desc(n))


```

#------------------------------------------------------------------------------
#Part 2: Cleaning Death Certificate Data
#------------------------------------------------------------------------------


## Check for duplicate names
```{r}

dc |> 
  count(name) |> 
  filter(n > 1)

```

### create separate list of duplicates
```{r}

dupes <- dc |> 
  count(name) |> 
  filter(n > 1) |> 
  select(name) |> 
  mutate(name = str_squish(str_replace_all(name,'\\.',''))) |> 
  pull(name)

```

### mark duplicates
```{r}

dc <- dc |> 
  mutate(dupes = case_when(
    name %in% dupes ~ "x")
  ) |> 
  arrange(personal_id) |> 
  select(dupes, everything()) 


```


**Meet with students to discuss duplicates**


# Causes of death
I am combining the cause of death from different columns unless there was a mistake
From the death certificate, we are counting all elements of the cause of death for a "causes of death" metric: Immediate cause of death; Intermediate (or antecedent) cause; Underlying cause of death. 
```{r}

dc <- dc |> 
  mutate(cause_death_update = paste(cause_of_death, if_cause_of_death_different_put_that_here, sep = "; ")) |> 
  mutate(cause_death_update = str_replace_all(cause_death_update, "NA", "")) |> 
  #removing pesky leading, trailing semicolons
  mutate(cause_death_update = str_remove_all(cause_death_update, "^;+|;+$"))    |> 
  mutate(cause_death_update = str_remove_all(cause_death_update, "^[;\\s]+|[;\\s]+$")) |> 
  mutate(cause_death_update = str_replace_all (cause_death_update, "blank", "")) |> 
  separate(col = cause_death_update, into = c("death_cause1", "death_cause2", "death_cause3"), sep = ";|and|And|or|&|/")

```

# Extract, tabulate death case
```{r}

death_categories <- dc |>
  pivot_longer(cols = c(death_cause1, death_cause2, death_cause3), 
               names_to = "cause_type", 
               values_to = "cause") |>
  count(cause) |>
  arrange(desc(n))
  

```

#Categorize
Split up into groups. Each group will tackle various diseases.
Coordinate so you don't overlap.

```{r}

death_categories <- death_categories |> 
    mutate(cause = str_remove_all(cause, "[*]")) |>
    mutate(cause = str_squish(cause)) |>
  mutate(category = case_when(
        cause=="Heart failure" ~ "heart_failure",
        cause=="And Pneumonia & Heart Failure" ~ "heart_failure",
        cause=="Heart Failure" ~ "heart_failure",
        cause=="Immediate: 2 heart failure" ~ "heart_failure",
        cause=="Immediate: Heart failure" ~ "heart_failure",
        cause=="Pneumonia and Heart failure" ~ "heart_failure",
        cause=="Pulmonary Tuberculosis" ~ "tuberculosis",
        cause=="Lobar Pneumonia" ~ "pneumonia",
        cause=="Catarrhal Pneumonia" ~ "pneumonia",
        cause=="Pneumonia (Double so called)" ~ "pneumonia",
        cause=="Pneumonia (unsure second w" ~ "pneumonia",
        cause=="Pneumonia Lab" ~ "pneumonia",
        cause=="Pneumonia phthisis" ~ "pneumonia",
        cause=="Primary Pneumonia" ~"pneumonia",
        cause=="Primary: Pneumonia " ~ "pneumonia",
        cause=='"Pneumonic Phthisis", Heart Failure' ~ "pneumonia",
        cause==" pneumonia" ~ "pneumonia", 
        cause==" Pneumonia" ~ "pneumonia",
        cause==" Pnuemonia" ~ "pneumonia", 
        cause=="Pneumonia " ~ "pneumonia", 
        cause==" Pneumonia " ~ "pneumonia", 
        cause=="Pneumonia" ~ "pneumonia",
        cause==" Asthenia" ~ "asthenia",
        cause==" Immediate: Asthenia" ~ "asthenia",
        cause==" asthenia" ~ "asthenia",
        cause=="Asthenia (27)" ~ "asthenia",
        cause=="Don't know, Stomack inflama   lungs congested" ~ "lung_disease",
        cause=="Oedema of Lungs" ~ "lung_disease",
        cause=="Hem" ~ "hemorrhage",
        cause=="rhage" ~ "hemorrhage",
        cause=="Hemm" ~ "hemorrhage",
        cause=="y Hemm" ~ "hemorrhage",
        cause=="wound of groin by hog, hem" ~ "hemorrhage",
        cause=="Influenza" ~ "Influenza",
        ## An online search shows oedema of lungs might be more closely associated with heart issues though
        cause=="rage of Lungs" ~ "lung_disease",
        TRUE ~ NA_character_
         ))
```

