---
title: "Bigrams for Cheltenham Research"
author: "Rob Wells"
date: '2025-11-24'
output: html_document
---

#Name: Nicole Ramos, Raphael Romero Ruiz, and Bex Heimbrock

# Jour 772 Fall 2025:

In this exercise, you will import the Article Data Extraction results and create bigrams from the comments
The source is Cheltenham Article Data Extraction (Responses), 184 entries as of Nov 24.

```{r}
#load tidyverse, tidytext, rio,, janitor libraries

library(tidyverse)
library(tidytext)
library(rio)
library(janitor)
library(googlesheets4)

```

# Load data - pull directly from spreadsheet
```{r}
#renamed important details to details for simplicity

library(googlesheets4)

gs4_deauth()  
news <- read_sheet("https://docs.google.com/spreadsheets/d/1XFRMzN-xK1uQcvaEIOV7wuEPH6lYrvVpvgE3w6ngKPU/edit?usp=sharing", sheet = "Form Responses 1") |> 
     clean_names() |> 
  rename(details = important_details_from_article_about_the_victim_and_the_conditions_at_cheltenham_any_direct_quotes_need_to_have_quotation_marks)

```

# Tokenize one column

```{r}

stories <- str_replace_all(news$details, "- ", "")
stories_df <- tibble(stories,)

# unnest includes lower, punct removal

stories_tokenized <- stories_df %>%
  unnest_tokens(word,stories)

stories_tokenized
#A list of the common words
```

#List of top words, cleaned with stop words removed
```{r}

data(stop_words)

stories_tokenized <- stories_tokenized %>%
  anti_join(stop_words, by = c("word" = "word")) %>%
  filter(word != "temp_file") %>%
  filter(word != "stories_corpus") %>%
  filter(!grepl('[0-9]', word))

# Word Count of top words

word_count <- stories_tokenized %>%
  count(word, sort=TRUE)


```

# Bigrams without stop words

```{r}
bigrams <- stories_df %>%
  unnest_tokens(bigram, stories, token="ngrams", n=2) |> 
  separate(bigram, c("word1", "word2"), sep = " ")

count_bigrams <- bigrams %>%
  count(word1, word2, sort = TRUE)


```

# Bigrams with stop words

```{r}
bigrams_clean <-stories_df %>%
  unnest_tokens(bigram, stories, token="ngrams", n=2) |> 
  separate(bigram, c("word1", "word2"), sep = " ") |> 
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word) |> 
  filter(!is.na(word1) | !is.na(word2))

count_clean_bigrams <- bigrams_clean  %>%
  count(word1, word2, sort = TRUE)


```

# Next steps: Categorize the material
Category "origins" = how the boys arrived -- sentenced, **committed** etc
"labor" - working on a farm, broom factory anything involving their work
"boy offenses" = specifies crimes the boys committed to get sentenced to Cheltenham
"investigations" = grand jury or newspaper investigations of Cheltenham - critical reviews
"abuse" = details of abuse, whippings, pubishment, no shoes, 
"medical" = discussing medical conditions
"praise" = any discussion that praises Cheltenham operations

#Bex, Nicole, and Raphael's additions start at line 144 in code
#NEW CATEGORIES
"escape" - escape, escaped, ran away, runaway, fled, runaways
"leadership" - Pyles, leadership, Dorsey, guards, guard, superintendent, teacher, teachers 
  
# NEW WORDS ADDED TO EXISTING CATEGORIES 
"abuse" - whip,whipped, whipping, punish, punishment, cruel, cruelty,  torture, tortured, beat, flogging, club, clubbed, clubbing
"neglect" - starv, malnourish, malnourished, unsanitary, filth, filthy 
"medical" - sick, illness, ill, hospital, doctor, physician, injury, injured, wound, disease, fever 
"death" - died, death, deceased, fatal, killed 
"origins" - sends, sent to, orphan, orphans
"boy_offenses" - charged, incorrigible, incorrigibility, indicted 
"investigation" - committee 
"labor" - parole, paroled 


Examine the bigrams and top words and on the categories below


```{r}
news <- news |> 
  mutate(details_clean = str_squish(str_remove_all(details, "[[:punct:]]"))) |> 
  mutate(category = paste(
    if_else(str_detect(details_clean, "(?i)sentenced"), "origins", NA_character_),
    if_else(str_detect(details_clean, "(?i)committed"), "origins", NA_character_),
    if_else(str_detect(details_clean, "(?i)labor"), "labor", NA_character_),
    if_else(str_detect(details_clean, "(?i)farm"), "labor", NA_character_),
    if_else(str_detect(details_clean, "(?i)broom"), "labor", NA_character_),
    if_else(str_detect(details_clean, "(?i)larceny"), "boy_offenses", NA_character_),
    if_else(str_detect(details_clean, "(?i)conditions"), "treatment", NA_character_),
    if_else(str_detect(details_clean, "(?i)jury"), "investigations", NA_character_),
    if_else(str_detect(details_clean, "(?i)beaten"), "abuse", NA_character_),
     # NEW PATTERNS ADDED BY NICOLE, BEX, RAPHAEL:
    if_else(str_detect(details_clean, "(?i)whip|(?i)whipped|(?i)whipping"), "abuse", NA_character_),
    if_else(str_detect(details_clean, "(?i)punish|(?i)punishment"), "abuse", NA_character_),
    if_else(str_detect(details_clean, "(?i)cruel|(?i)cruelty"), "abuse", NA_character_),
    if_else(str_detect(details_clean, "(?i)torture|(?i)tortured"), "abuse", NA_character_),
    if_else(str_detect(details_clean, "(?i)neglect|(?i)neglected"), "neglect", NA_character_),
    if_else(str_detect(details_clean, "(?i)starv|(?i)malnourish|(?i)malnourished"), "neglect", NA_character_),
    if_else(str_detect(details_clean, "(?i)unsanitary|(?i)filth|(?i)filthy"), "neglect", NA_character_),
    if_else(str_detect(details_clean, "(?i)sick|(?i)illness|(?i)ill"), "medical", NA_character_),
    if_else(str_detect(details_clean, "(?i)hospital|(?i)doctor|(?i)physician"), "medical", NA_character_),
    if_else(str_detect(details_clean, "(?i)injury|(?i)injured|(?i)wound"), "medical", NA_character_),
    if_else(str_detect(details_clean, "(?i)disease|(?i)fever"), "medical", NA_character_),
    if_else(str_detect(details_clean, "(?i)escape|(?i)escaped"), "escape", NA_character_),
    if_else(str_detect(details_clean, "(?i)died|(?i)death|(?i)deceased"), "death", NA_character_),
    if_else(str_detect(details_clean, "(?i)fatal|(?i)killed"), "death", NA_character_),
    if_else(str_detect(details_clean, "(?i)sends|(?i)sent to"), "origins", NA_character_),
    if_else(str_detect(details_clean, "(?i)charged"), "boy_offenses", NA_character_),
    if_else(str_detect(details_clean, "(?i)incorrigible|(?i)incorrigibility"), "boy_offenses", NA_character_),
    if_else(str_detect(details_clean, "(?i)beat"), "abuse", NA_character_),
    if_else(str_detect(details_clean, "(?i)committee"), "investigation", NA_character_),
    if_else(str_detect(details_clean, "(?i)parole|(?i)paroled"), "labor", NA_character_),
    if_else(str_detect(details_clean, "(?i)indicted"), "boy_offenses", NA_character_),
    if_else(str_detect(details_clean, "(?i)flogging"), "abuse", NA_character_),
    if_else(str_detect(details_clean, "(?i)orphans|(?i)orphan"), "origins", NA_character_),
    if_else(str_detect(details_clean, "(?i)superintendent|(?i)guard|(?i)guards|(?i)teacher|(?i)teachers"), "leadership", NA_character_),
    if_else(str_detect(details_clean, "(?i)Pyles|(?i)leadership"), "leadership", NA_character_),
    if_else(str_detect(details_clean, "(?i)Dorsey"), "leadership", NA_character_),
    if_else(str_detect(details_clean, "(?i)club|(?i)clubbed|(?i)clubbing"), "abuse", NA_character_),
    if_else(str_detect(details_clean, "(?i)ran away|(?i)runaway|(?i) runaways|(?i)fled"), "escape", NA_character_),
    sep = "; "
  )) |> 
  mutate(category = str_remove_all(category, "NA; |; NA")) |> 
  mutate(category = na_if(category, "NA"))
```



# Next steps: Team up:
Akira, Brandyn
Nicole, Bex, Raphael
Sophie, David, Zaka

1) bigrams and word count on one of these three columns "elaborate_on_the_quality_of_the_journalism_sourcing_graphics_photos_interviews"; "notes_on_neglect_or_abuse"; "notes_anything_else_thats_noteworthy"     

## Tokenize "elaborate_on_the_quality_of_the_journalism_sourcing_graphics_photos_interviews"column
```{r}
journalism <- str_replace_all(news$elaborate_on_the_quality_of_the_journalism_sourcing_graphics_photos_interviews, "- ", "")
journalism_df <- tibble(journalism,)
# unnest includes lower, punct removal
journalism_tokenized <- journalism_df %>%
  unnest_tokens(word, journalism)
journalism_tokenized
```
## Word count for "elaborate_on_the_quality_of_the_journalism_sourcing_graphics_photos_interviews"column (cleaned with stop words removed)
```{r}
journalism_tokenized <- journalism_tokenized %>%
  anti_join(stop_words, by = c("word" = "word")) %>%
  filter(word != "temp_file") %>%
  filter(word != "stories_corpus") %>%
  filter(!grepl('[0-9]', word))
# Word Count of top words
journalism_word_count <- journalism_tokenized %>%
  count(word, sort=TRUE)
journalism_word_count
```
## Bigrams without stop words for  "elaborate_on_the_quality_of_the_journalism_sourcing_graphics_photos_interviews"column
```{r}
journalism_bigrams <- journalism_df %>%
  unnest_tokens(bigram, journalism, token="ngrams", n=2) |>
  separate(bigram, c("word1", "word2"), sep = " ")
count_journalism_bigrams <- journalism_bigrams %>%
  count(word1, word2, sort = TRUE)
count_journalism_bigrams
```
## Bigrams with stop words removed "elaborate_on_the_quality_of_the_journalism_sourcing_graphics_photos_interviews"column
```{r}
journalism_bigrams_clean <- journalism_df %>%
  unnest_tokens(bigram, journalism, token="ngrams", n=2) |>
  separate(bigram, c("word1", "word2"), sep = " ") |>
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word) |>
  filter(!is.na(word1) | !is.na(word2))
count_clean_journalism_bigrams <- journalism_bigrams_clean %>%
  count(word1, word2, sort = TRUE)
count_clean_journalism_bigrams
```
2) add to the category code above -- be specific on what you are adding so I can merge it into the master code block

#see above additions (line 112-125 for descriptions and 144 in code)
