
#Name: Nicole Ramos
#Date: 2025-09-25
#Assignment: Lab 03

---
title: "lab_03"
author: "Derek Willis/ Rob Wells"
date: "2025-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

library(tidyverse)
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows

payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=369008)
  payments <- payments |>
  clean_names()


# Load in the Maryland grants and loan data

md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv", guess_max=19482) 
  md_grants_loans <- md_grants_loans |>
  clean_names()

```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. 

Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1:
#If I did it right, I found 88 records. 

```{r}
pg_schools <- payments |> filter(str_detect(vendor_name, 'PRINCE GEORGES COUNTY PUBLIC SCHOOLS') | str_detect(vendor_name, 'BOARD OF EDUCATION OF PRINCE GEORGES COUNTY') | str_detect(vendor_name, 'PG COUNTY PUBLIC SCHOOL') | str_detect(vendor_name, 'BD OF EDUCATION OF PG CO') | str_detect(vendor_name, 'BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD') | str_detect(vendor_name, 'PRINCE GEORGES PUBLIC SCHOOLS') | str_detect(vendor_name, 'PRINCE GEORGES COUNTY SCHOOL') | str_detect(vendor_name, 'BOARD OF EDUCATION OF PRINCE GEORGE'))
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2:
#Fiscal Year 2009 had the most amount of payments: 8. Fiscal year 2024 had the highest payment at 1782519925. I am doing some research on PG County right now, and know they are undertaking massive rebuilds, which might be why there are fewer payments than 2009 but those that are paid are bigger.

```{r}
pg_schools %>% 
group_by(fiscal_year) %>% 
summarise(count = n(),total = sum(amount, na.rm=TRUE)) |>
  arrange(desc(total))
```

Q3. This question is in two parts. 

Part 1: Code: You're interested in examining the vendors paid for capital projects at the University of Maryland. Write code that does the following: 
--Filters payments data to capital projects at the University of Maryland
--Produce a list of vendors and the total amount they received. 


A3.
#Part 1: 

```{r}
capital_umd <- payments |>
  filter(agency_name == 'CAPITAL PROJECTS- UNIVERSITY OF MARYLAND')

capital_umd |>
  group_by(vendor_name, vendor_zip) |>
  summarise(total = sum(amount, na.rm=TRUE))
```

Part 2: Provide a written answer, not code.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? 
--Provide at least three examples of vendor names that are similar and should be combined into one. 
--Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

#Part 2: 

#We should keep capital_umd as a complete dataframe so that we can look for possible duplicate entries, ensuring they are in fact duplicates by also checking date, zip, and/or amount. There doesn't seem to be according to get_dupes. We should then look for vendor names to see if they are slightly different but represent the same vendor. The second part of the code block, grouping by vendor name and zip, can help this process by de-cluttering some of what we are looking at. 

#AYERS SAINT GROSS INC, A MORTON THOMAS & ASSOCIATES at zip 20852, AERSOL MONITORING & ANALYSIS INC, and COLIMORE THOEMKE ARCHITECTS INC are similar and should be combined. 

#It might be helpful to clean up the data using mutate and string commands to standarize the variables in the vendor_name column. Checking my notes from last week, str_remove might be a tool we could use too. We could also use str_replace_all maybe to take the spacing out. That could help in situations like STORAGELOGIC OF MARYLAND INC, which was separated from the STORAGE LOGIC OF MARYLAND INC entry. 

```{r}
capital_umd |> get_dupes()

capital_umd |> mutate(vendor_name = str_replace_all(vendor_name, " ", ""))

capital_umd |> mutate(vendor_name = str_remove(vendor_name, ","))
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. 

Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. 
Then create a total amount of money for each fiscal year. 

Write a sentence summarizing how stem cell research funding has changed over time. Be specific with the totals.

A4:
#According to the code I ran, stem cell research funding hit the multiple digit millions in the earliest and most recent fiscal years we have in the data set, 2009 and 2023. It hit a low in 2021 with only $4832225 in grants being awarded, which was likely due to the covid pandemic. It seems like there are two to three year periods where the subsequent years get awarded a similar amount of money, for example 2014 ($6785990) 2015 (6454590). Perhaps these were grants for multi-year research studies, predicated on the year before. Maybe more importantly, FY2016 and FY2017 didn't include any stem cell research grants.  

```{r}
stem_cell_research <- md_grants_loans|>
  mutate(description = str_to_upper(description)) |>
  filter(str_detect(description, "STEM CELL RESEARCH ")) |>
  group_by(fiscal_year)|>
  summarise(total = sum(amount, na.rm=TRUE))

```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5:
#My code identified 18 grantees, though it looks like some could be collectively standardized as University of Maryland, College Park. I could join the entries of UNIVERSITY OF MD,COLLEGE PARK and University of Maryland College Park by using str_upper to standardize the case. There is one entry for University of Maryland that I would want to know more about before deciding whether or not to also include it as well. GripBoost, which apparently is a sports glove  company, is odd, but upon further investigation of their website, they were founded by UMD chemical engineers. They could have gotten some sort of start up grant. 

```{r}
md_grants_loans |>
  mutate(zip5 = str_sub(zip_code, start=1L, end=5L)) |>
   mutate(zip5 = case_when(
    zip5 == "Vario" ~ NA,
    zip5 == "UB7 O" ~ NA,
    zip5 == "UB7 " ~ NA,
    .default = zip5))|>
  filter(zip5 == '20742') |>
  group_by(grantee)|>
  summarise(total = sum(amount, na.rm=TRUE))
```
