---
title: "Brandyn's Data Notebook"
author: "Brandyn Fragosa"
date: "2025-10-08"
output: html_document
---
###This R file will include a database that examines the rate of police shootings by both race and gender in America. The orginal database that this file in built on is from a Washington Post data that led to a 2020 article. We worked to recreate the same information in the article built off the same data. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
options(scipen=999)
library(tidyverse)
library(janitor)
library(lubridate)
```

```{r}
fatal_police_shootings_data <- read_csv("fatal-police-shootings-data.csv") %>% 
  clean_names() %>% 
  mutate(date_new = ymd(date))
```
**Brandyn**
```{r}
# Q2. Number of fatal police shootings by gender since 2015 to 09/04/2020 (publication of the article.

shootings_by_gender <- fatal_police_shootings_data %>% 
  filter(date_new < "2020-09-04") %>% 
  filter(
    str_detect(gender, "male")|
    str_detect(gender, "female")
)%>% 
group_by(gender) %>%
summarise(count = n())
```

```{r}
# Q2. Men and woman shot by police since 2015 who had mental illnesses.

Mental_illness_related_by_gender <- fatal_police_shootings_data %>% 
  filter(date_new < "2020-09-04") %>% 
  filter((str_detect(gender, "male")|str_detect(gender, "female"))&
  str_detect(was_mental_illness_related, "TRUE"))%>% 
group_by(gender) %>%
summarise(count = n())
```

```{r}
# Q2. Percentage of men and woman shot by police since 2015 who had mental illnesses.

total_women_shot <- fatal_police_shootings_data %>% 
    filter(date_new < "2020-09-04") %>% 
    filter(str_detect(gender, "female")) %>% 
  group_by(gender) %>%
  summarise(total = n())

women_with_mental_illness <- fatal_police_shootings_data %>% 
    filter(date_new < "2020-09-04") %>% 
    filter(
      str_detect(gender, "female"),
      str_detect(was_mental_illness_related, "TRUE"))%>% 
    group_by(gender) %>% 
    summarise(total = n ())

#I was having so much trouble grabbing the total from the two data sets and could not figure it out, I tried using chatgpt to explain whats wrong with my code but nothing would give me the values I needed.

total_women_shot <- 248
women_with_mental_illness <- 83
Pct_with_mental_illness <- (women_with_mental_illness / total_women_shot) * 100
```

**Akira Kyles**
```{r}
#By race, 147 of the women killed were White, 48 Black and 29 Hispanic. Five were Native American, four were Asian and three were other races.

fatal_police_shootings_data |> 
  filter(date < "2020-09-01" &
           gender== "female") |> 
  group_by(race) |> 
  summarise(count=n()) 
```
##By percentages, men shot by police were less often White and more often Black or Hispanic. 

##My data shows this statement as false and it was actually the complete opposite and white men are actually shot more than than Black men by percentage.    
```{r}


fatal_police_shootings_data |> 
  filter(date < "2020-09-03" &
           gender== "male" & 
           armed_with== "gun" & 
           threat_type== "shoot") |> 
  group_by(race) |> 
  summarise(count = n()) |> 
  mutate(percentage = (count / sum(count)) * 100)
```
### For both genders, a gun was the most common weapon. Of men killed, 57 percent were armed with a gun and of women, 44 percent.
```{r}
deaths |> 
  filter(date > "2020-09-03" &
           gender== "female" ) |>
  group_by(armed_with) |> 
  summarise(count = n()) |> 
  mutate(percentage = (count / sum(count)) * 100) |> 
  arrange(desc(percentage))

```

```{r}
deaths |> 
  filter(date > "2020-09-03" &
           gender== "male" ) |>
  group_by(armed_with) |> 
  summarise(count = n()) |> 
  mutate(percentage = (count / sum(count)) * 100) |> 
  arrange(desc(percentage))
```
###The data, although very slightly higher in percentage by minimal margins, those armed with guns is the highests for both genders. In the Washington Post data, 109 women were reported to be armed with a gun but I found 105 and my percentage increased by 4% compared to Washington Post results. The margins for the male data was much closer with 59% compared to the Washington Post's 57%

**Nicole Ramos**

### For 5A, I took the dataset and filtered for two factors: dates before Sept. 3, 2020 and female in the gender column. I grouped the newly filtered data by the column "armed_with," in order to see the counts of females by what weapons they were (or weren't) carrying when killed by police.

```{r}
#5A. Since 2015, police have killed 26 unarmed women, including Taylor.

fatal_police_shootings_data |>
  filter(date_new < "2020-09-03" & 
           gender== "female") |>
  group_by(armed_with) |>
  summarise(count = n())
```

### For 5B, I retained the two previous filters but added one more: unarmed in the "armed_with" column. I then grouped by race and summarized, so I could see the counts of unarmed women by race. 

### 5C's statement doesn't specify whether the Black and White women they are referring to are armed or unarmed. I ran code below, removing the unarmed column filter and grouping by race, to see if that data came from women generally. It did not appear so, as White women fatally shot by police would otherwise be about three times as many as Black women. 5C does appear true when referring to the unarmed female counts by race.


```{r}
#5B.Of those, 14 were White, seven were Black, four were Hispanic and one’s race was unknown.

#5C.While about twice as many White women were shot and killed as Black women...

fatal_police_shootings_data |>
  filter(date_new < "2020-09-03" & 
           gender== "female" &
           armed_with == "unarmed") |>
  group_by(race) |>
  summarise(count = n())
```

```{r}
#5C.The second statistic quoted above seems to refer to unarmed White women in comparison to unarmed Black women, because White women fatally shot by police would otherwise be about three times as many as Black women. 

fatal_police_shootings_data |>
  filter(date_new < "2020-09-03" & 
           gender== "female") |>
  group_by(race) |>
  summarise(count = n())
```

  
# 5D. White women account for five times the Black female population.

### Given that The Washington Post's GitHub does not contain any csv files for total U.S. population, the statistic could come from 2010 census data (https://www.census.gov/library/visualizations/interactive/race-and-ethnicity-in-the-united-state-2010-and-2020-census.html). In 2010, the population identifying as White alone was 61.6%, which is five times more than the population identifying as Black, 12.4%. These numbers, though, are not filtered out for specific gender.  