---
title: "lab_08_update"
author: "Derek Willis/Rob Wells"
date: "2025-10-27"
output: html_document
---

#Name: Nicole Ramos 
#Assignment: Lab_08
#Date: 10-29-2025


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. We'll be making some charts, working with dates and retrieving Census data.

```{r}
options(scipen=999)
library(tidyverse)
library(tidycensus)
library(lubridate)
library(ggthemes)
```

## Load and modify data

**Task** Create a codeblock and load the following data from the data folder:

-   Prince George's County 911 Overdose Calls

```{r}
pgco_911_calls <- read_csv("data/prince_georges_2022_overdoses.csv") |>
   mutate(call_date=as.Date(datetime))

pgco_911_calls <- pgco_911_calls |>
 mutate(week = floor_date(call_date, "week"))  |>
  mutate(month = month(call_date, label=TRUE))
```


You will need to create columns for the date, week and month based on the existing `datetime` column.

## Questions

**Q1.** Which month saw the greatest percentage of total calls? Create a dataframe that calculates the percentage of all calls that each month's calls represents. Make a bar chart from that dataframe. Your bar chart must have:

-   A clear title that states the main idea/finding
-   Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
-   Readable bars - the values shouldn't be overlapping


**A1.**

```{r}
pgco_911_calls_month<- pgco_911_calls |> 
  group_by(month) |> 
  summarize(total_calls = n()) |> 
  mutate(pct_total_calls= (total_calls/sum(total_calls))*100)
```

```{r}
pgco_911_calls_month |>
 ggplot() +
  geom_bar(aes(x=reorder(month,pct_total_calls), weight=pct_total_calls)) +
  labs(
    title="911 Calls Peak Around the Holiday Season",
    x = "Month",
    y = "Percent of Total Calls",
    caption = "Source: Prince George's County EMS"
    
  )
```

**A1-written answer: Which month saw the greatest percentage of total calls? Write that here:**

#December was the month with the greatest percentage of total calls, 10.09%. 

**Q2.** Let's visualize this data on a weekly basis using a line chart. As in Q1, generate a dataframe with the total number of calls for each week, and then create a line chart to show the distribution of calls over time. Your line chart must have:

-   A clear title that states the main idea/finding
-   Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
-   Readable labels


**A2.**

```{r}
pgco_911_calls_week<- pgco_911_calls |> 
  group_by(week) |> 
  summarize(total_calls = n()) |>
  arrange(desc(total_calls))
```


```{r}
pgco_911_calls_week |>
  ggplot() + 
  geom_line(aes(x=week, y=total_calls)) + 
  scale_x_date(date_breaks = "4 week", date_labels = "%b %d") +
    theme(
    axis.text.x = element_text(angle = 45,  hjust=1)) +
    labs(
    title="911 Calls Spike the Week of Fourth of July",
    x = "Week",
    y = "Total Calls",
    caption = "Source: Prince George's County EMS")
```


**A2--written answer:Describe the pattern of results; you may want to look at the data to dig into particular outliers.Write that here:**

#911 calls are higher during the week of Fourth of July than any other week. It's also interesting that the week between Christmas and New Year's is significantly lower than the rest of the year, by 10 calls when compared to the next lowest week.


**Q3.** A scatterplot is a type of chart that helps us see relationships between two variables. One variable goes on the x axis, the other on the y axis. For each row/observation in our data, a scatterplot puts a circle (or a "point") where the two variables intersect on a grid.

Statisticians use scatterplots to show graphically whether one variable is correlated -- related, in a statistical sense -- with another variable. A classic example is the [relationship between ice cream sales and temperature](https://www.mathsisfun.com/data/scatter-xy-plots.html). The scatterplot below -- press play to load the image -- shows that relationship, that an increase in temperature is associated with an increase in ice cream sales. When it's 12C, sales are 200 dollars, and when it's hotter, 25C, sales are 600 dollars.

```{r}
knitr::include_graphics("https://www.mathsisfun.com/data/images/scatter-ice-cream1.svg")
```

We're going to use a scatterplot a little differently, to get a visual sense of two key variables:

Our question is: does the median income in a zip code have any relationship to the number of overdose 911 calls in that zip code?

To answer this question, do the following:

**Q3 - 1.**  Generate a dataframe with the number of 911 calls for each zip code.
```{r}
pgco_911_calls_zip<- pgco_911_calls |> 
  group_by(zipcode) |> 
  summarize(total_calls = n()) |>
  arrange(desc(total_calls))
```

**Q3 -2.**  Get data from the Census Bureau showing median household income for Maryland zip codes.

```{r echo=FALSE}
census_api_key("", install=FALSE)
```

```{r}
acs5 <- load_variables(2022, "acs5", cache = TRUE)
View(acs5)
```

```{r}
md_zip <- get_acs(geography="zcta", 
                  variables = "B19013_001", 
                  year=2022) |>
    filter(GEOID >= "20601" & GEOID <= "21930")
```

**Q3 -3.**  Join those two dataframes on their zip code columns, starting with the 911 calls dataframe.
```{r}
md_zip_with_income <- pgco_911_calls_zip |> inner_join(md_zip, by=c('zipcode' = 'GEOID')) |> rename(median_income = estimate)
```

**Q3 -4.**  Make a scatterplot showing the total calls and median income. I didn't show you how to do this, so look it up! Googling "ggplot scatterplot" is a good start. Give it an appropriate title, source, and x and y axis titles. 
```{r}
ggplot(data = md_zip_with_income, aes(x = median_income, y = total_calls)) +
      geom_point() +
   labs(
    title="Households Under $100K are Highest 911 Callers",
    x = "Median Income",
    y = "Total Calls",
    caption = "Source: 2018-2022 5-year ACS & Prince George's County EMS")

```

**Bonus question:** Take the scatterplot above and add a label for each point that shows the zip code using geom_text() - see some examples of its use at <https://ggplot2.tidyverse.org/reference/geom_text.html#ref-examples>. Try to make the names as easy to read as possible by avoiding overlap.

```{r}
ggplot(data = md_zip_with_income, aes(x = median_income, y = total_calls, label=zipcode)) +
  geom_label(check_overlap = TRUE, hjust = 1, nudge_x = .5)
   labs(
    title="Households Under $100K are Highest 911 Callers",
    x = "Median Income",
    y = "Total Calls",
    caption = "Source: 2018-2022 5-year ACS & Prince George's County EMS")
```


**Q4**.  In the answer space below, describe what you see and answer the questions posed above. In a general sense, what do you think this means? Feel free to consider the actual raw values: how would you report out the main point(s) of this chart?

**A4: Write your answer below.**

#I see, as described in my title, that the highest 911 callers are people whose median income is less than $100k a year. 

#I think it could also helpful to visualize this on a map, perhaps one that shows variations of colors based on amount of 911 calls. 

#It is important to report that the lowest median income points we have are not the highest 911 callers, contrary to what people might assume. I would also report more information on the demographic and makeup of the zip codes with the highest 911 calls. 

#It might also be interesting to figure out the average calls for Prince George's County and compare it to other years. 