
Name: Nicole Ramos
Assignment: lab_01
Date: 2025-09-10

---
title: "lab_01"
author: "Derek Willis / Rob Wells"
date: "2025-09-10"
output: html_document
---

PUT YOUR NAME AND DATE ON THIS ASSIGNMENT
#above


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you? 

**A1**
#The department with the highest average is the Business and Management department. I am actually suprised English ranks as 4th. 

```{r}
umd_courses |>
  group_by(department) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    mean_seats = mean(seats),
  ) |>
  arrange(desc(total_seats))
```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your result should sort to show the greatest number of classes first. If several classes are tied for the most offered, describe how many classes are ranked in the top category for offerings.

**A2**
#There are 145 individual classes offered 29 times. 

```{r}
umd_courses |>
  group_by(id) |>
  summarize(count_classes = n()) %>% 
  arrange(desc(count_classes))
```


**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most? HINT: To filter out blank and missing values from a column, drop_na() is your friend.

**A3**
#If I did this right, Keith Yearwood from the Geographical Sciences department taught the most classes at 79. However, when I ran the code chunk below, there are 97 classes in the Biochemistry department with an instructor TBA.
```{r}
umd_courses |>
  group_by(instructors, department) |>
    summarise(
    count_classes = n()
  ) |> drop_na() |>
  arrange(instructors, department) |>
  arrange(desc(count_classes))

```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? 

**A4**
#My first instinct when I saw the code chunk output was to filter out any instances of "Instructor: TBA." I am not sure how I would filter out a character based on grouping, or even filter out in general. I think I have a basic understanding of filtering (grepl) based on a word (i.e. Wall Street Journal example), but I don't think I know how to filter out a character. I think I got the right answer, though we discussed using str_split to split the column with multiple instructor names. 

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? 

**A5**
# 1.How many courses contain "data" in their description? 
#There are 961 courses that have data in their description.I think there might be another way to do this, that doesn't require having to make a separate table altogether. What I did that felt easiest was create a subset, filtering any description of courses that contained the word "data." I do think it's interesting that so few courses actually contain data elements. 

# 2.What does the subset of courses look like that have 0 sections? 
#I obviously played around with this until I got it to do what I wanted it to. When I initially created a subset based on "0", it was returning courses that had, for example, 10. Instead of using the grepl command, I used filter, and set the filter to classes that equalled zero. I learned in the process == is equal to, not =. I do think there are so many zero section courses. I wonder how many students have to sign up for a course to be offered. 

```{r}

Datacourses <-subset(umd_courses, (grepl("?data", description)))

```

```{r}
Nocourses <-subset(umd_courses, (grepl("?0", sections)))
```

```{r}
Nocourses2 <- umd_courses %>% 
filter(sections== 0)
```

