
#Name: Nicole Ramos
#Assignment: lab_07
#Date: 2025-10-24

---
title: "lab_07"
author: "Derek Willis/Rob Wells"
date: "2025-10-21"
output: html_document
---


Lab07 - API Keys and Tidycensus


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
Sys.getenv("R_PROFILE_USER")
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidycensus)
library(tidyverse)
```

## Security and API keys

You'll need your Census API Key. Here's how to do that securely:

r_environ is a hidden file that can securely store API keys
```{r}
usethis::edit_r_environ()
# a new text file opens in RStudio
```

1. Add your API keys in the r_environ file
CENSUS_API_KEY=XXXXX
OPENAI_API_KEY=XXXX
etc....

2. Save the r_environ() file and RESTART R

# Hide your API Key from GitHub

```{r}
# Check if .gitignore exists in your project
file.exists(".gitignore")
```

Assuming you don't have a .gitignore file, then create one
1. RStudio:
- File → New File → Text File

2. ## Add These Lines to Your `.gitignore`
# R files
.Rhistory
.RData
.Rproj.user/
.Renviron
.Rprofile

# Environment files
.env
*.env

# API keys and secrets
*secret*
*password*
*api_key*

3. save the file a .gitignore in the root directory of your project
yes, you have to precede the file with a dot: .gitignore

4. Veryify it is working

```{r}
system("git status")
```


## Setup Census API

You'll need your Census API Key:

**Task** set your API Key if you haven't already. You won't need to install or overwrite it.
```{r echo=FALSE}
census_api_key("", install=FALSE)
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2023 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2023 5-year ACS (which is called "acs5") and save them to a dataframe

```{r}
acs5 <- load_variables(2023, "acs5", cache = TRUE)
View(acs5)
```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2023 5-year estimates? Hint: look at the variables for ACS in 2023 and find the variable for median age. 

**A1**

#The jurisdiction with the lowest median age appears to be Block Group 3; Census Tract 4514.02; Baltimore County; Maryland with a median age of 5.7, verified with Census Reporter: [https://censusreporter.org/profiles/15000US240054514023-bg-3-tract-451402-baltimore-md/]. Based on the results for Q2, Baltimore City is the county with the lowest median age, 36.1.

```{r}
md <- get_acs(geography = "block group",
              variables = c(medage = "B01002_001"),
              state = "MD",
              year = 2023)

md <- md |> arrange(estimate)

md
```

**Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe created in Q1 to answer this, but you must write code to display only those counties with a median age of at least 50. 

**A2**

#Talbot and Worcester are the only counties with a median age at least or greater than 50. 

```{r}
md_county <- get_acs(geography = "county",
              variables = c(medage = "B01002_001"),
              state = "MD",
              year = 2023)
md_county |>
  filter(estimate >= '50')
```

**Q3** We're interested in knowing more about Montgomery County zip codes where overdose calls are coming from and the demographic characteristics they possess. In particular, we want to calculate a rate of calls per 1,000 population. To do this, you'll need to:

1.  Load the Montgomery County 911 overdose calls data.
```{r}
mont_county_911 <- read_csv("data/montgomery_2022_overdoses.csv") |>
  mutate(zip = as.character(zip))
```

2.  Create a new dataframe with the total number of calls for each zip code.
```{r}
mont_county_911_zip <- mont_county_911 |>
  group_by(zip) |>
  summarize(total_calls=n())
```

3.  Find the ACS variable pertaining to the total 18 and up population (there are several choices here; you want the simplest one). Then get the ACS data from all Maryland zip codes using that variable - you will need to specify an earlier year to get the data.
```{r}
md_zip <- get_acs(geography="zcta", variables = "B01001_001", year=2022)|>
  filter(GEOID >= "20601" & GEOID <= "21930")
```

4.  Join the ACS data to the Montgomery County zip code totals dataframe and rename the estimate column to a meaningful name you've given the ACS variable.
```{r}
mont_county_911_acs <- mont_county_911 |> left_join(md_zip, by=c('zip'='GEOID')) |>
  rename(total_18_up = estimate)
```


#rsw comment:
I did an inner join
```{r}
montco_zip_with_population <- montco_totals |> inner_join(md_zcta_population, by=c('zip' = 'GEOID')) |> rename(population = estimate)

```


5.  Add a column to the joined dataframe that calculates a rate of calls per 1,000 population.
```{r}
mont_county_911_acs |>
  group_by(zip, total_18_up) |>
  summarize(total_calls=n()) |>
  mutate(calls_per_1k = (total_calls/total_18_up)*1000) |>
  arrange(desc(calls_per_1k))
```


#rsw comment: see my code - i didnt use the group_by
```{r}
montco_zip_with_population <- montco_zip_with_population |> mutate(rate_per_1k = (calls/population)*1000) |> arrange(desc(rate_per_1k))

```


Which zip code has the highest rate of calls? Which has the lowest?

**A3**
#Zip code 20707 has the lowest rate of calls, at 0.028 per 1,000.
#Zip code 20889 has the highest rate of calls, at 6.329 per 1,000, however, because the populaton is under 1,000, I might also consider 20877. 


#rsw comment: 20877 is the highest, 20707 is the lowest.
```{r}

```

**Q4** Using [CensusReporter.org](https://censusreporter.org/), examine the zip codes with the highest and lowest rates and compare and contrast their demographic profiles. Is there a story here?

**A4**

**Write your answer here:**
#I focused on 20707 and 20877, as both have populations over 18 more than 1,000. 20707, with the lowest crime rate, has a median income close to $20,000 more than 20877, the second highest crime rate. 20877 has a poverty rate of 11.7%, which Census Reporter summarizes as 20% higher than the [average?] rate in Maryland, whereas 20707's poverty rate falls below that of Maryland. Close 50% of the population in 20877 is Hispanic, whereas 45% of the population is Black in 20707. 

#I do think there is a story here, with respect to the the rate of overdoses for zip codes that are majority Hispanic and falling below the poverty line. What stressor(s) are leading to this? Have rates increased from 2021-2022? What about more recent data? What resources are or aren't accessible?
